# Introduction

In this vignette, we will explore the functions in the LatticeKrig package and show examples of how they can be used to solve problems. The LatticeKrig (LK) model is an example of the spatial statistics method known as kriging, adapted to large data sets.

## What is kriging?

  By spatial data, we mean the data contains the observed variable and
its location, the variable depends on the location, and  observations that are closer  are  more
similar values than those farther apart.

Two  quotes by Waldo Tobler sum this up nicely:
\begin{quote}
\textit{Everything is related to everything else,
           but near things are more related than distant things} 
  
\textit{The phenomenon external to an area of interest affects what goes on inside}.
\end{quote}  
In the methods of this package, we quantify this relatedness  of "near things"",  statistically using correlation.

Kriging (named for South African statistician Daniel Krige) is a method for making predictions from a
spatial data set first developed for mineral exploration.
As such, kriging can be applied to a variety of important
data sets, from geological data to atmospheric data. Moreover, the method 
generalizes to the coordinates of a surface that need not be geographic or 2D/3D physical locations.
The classic method uses a model for the covariance (correlation) of the observations as a function 
of their distance of separation.
The Kriging predictions are weighted sums of the observations, where the weights depend on the
covariances and are different at every location. 

The standard spatial model for Kriging relates the observation to a sum of three components:
a \textit{fixed part} typically a polynomial function of the locations (and covariates,
if provided),
a \textit{spatial process}, and \textit{measurement error}. This is a more practical model for data
analysis but involves more complicated formulae for the Kriging weights. The LatticeKrig (LK) model
like most spatial methods, decomposes the data into these three components.

## The LatticeKrig model

The key feature of LatticeKrig is that we model a spatial process as the sum of radial basis functions (functions that are symmetric around their center and are 0 for far away points) scaled by coefficients, which we assume are correlated. The smooth basis functions and correlated coefficients create a smooth function representation for the spatial process, and the structure of the basis functions and covariance has some flexibility so you can change the structure to make a more reasonable model for a certain problem. The linear polynomial in the locations and covariates is determined using generalized least squares, following the standard approach of Universal Kriging. To approximate the spatial process, we then fit the basis functions to the residuals from the linear model. In terms of linear algebra, the model is
$$\mathbf y = X \mathbf d + \Phi \mathbf c + \mathbf e$$
where $\mathbf y$ is the vector of variable measurements, $X$ is the matrix of locations and covariates, $\mathbf d$ is the vector of coefficients for the linear model, $\Phi$ is the matrix of basis functions evaluated at the data points, $\mathbf c$ is the vector of coefficients for each basis function, and $\mathbf e$ is the measurement error. If we let $\mathbf g$ represent the true values of the variable (without measurement error $\mathbf e$), we can unroll the matrix multiplications into sums and get the form
$$ g(\mathbf s) = \sum_{k=1}^n \phi_k(\mathbf s) \hat {\mathbf d}_k + \sum_{k=1}^m \psi_k (\mathbf s) \hat{\mathbf c}_k$$
We show the derivations of the equations for $\mathbf c$ and $\mathbf d$ in Appendix A, and show how all of these calculations are done in Appendix C.

The package is named LatticeKrig because of the placement of the basis functions: they are equally spaced in all dimensions on a lattice. We can also consider multiple different lattice sizes simultaneously to better capture different levels of resolution; by default, each additional level has half as much space between the basis functions in each dimension.
The following plot shows the basis functions for each of the three default levels.

But before we start with code here are the R, fields package, and LatticeKrig package versions
used throughout this vignette. Use these to check your versions if you get errors and we will try to make this vignette compatible with latter versions.   

```{r versions, echo=TRUE}
R.Version()$version.string
packageVersion( "fields")
packageVersion( "LatticeKrig")
```

```{r levelsPlot, echo=FALSE}

phi <- function(d) {
  return(1/3 * (1-d)^6 * (35*d^2 + 18*d + 3) * (d < 1))
}
overlap <- 2.5
gridPoints <- seq(0, 10, length=1000)
centers1 <- seq(0,10, length=11)
centers2 <- seq(0,10, length=21)
centers3 <- seq(0,10, length=41)
dists1 <- rdist(gridPoints, centers1)
dists2 <- rdist(gridPoints, centers2)
dists3 <- rdist(gridPoints, centers3)
values1 <- phi(dists1 / overlap)
values2 <- 1/2 * phi(2 * dists2 / overlap)
values3 <- 1/4 * phi(4 * dists3 / overlap)
matplot(x = gridPoints, cbind(values1, values2, values3), type="l", xlab="Location",
        ylab = "Basis function value", main="Basis Functions for 3 Levels", lty = 1,
        col = c(rep("black", 11), rep("red", 21), rep("green", 41)))
legend("topright", c("First Level", "Second Level", "Third Level"),
       col=c("black", "red", "green"), lty=1, bg="white")
#lines(values[,7], x=gridPoints, type="l", col="black", lwd=3)
```

Using three levels in practice, similar to those in the plot above, we can add the contributions from each level to capture spatial patterns more precisely at different scales.

\textbf{First Level:} Captures broad, large-scale trends in the data with basis functions further apart that cover larger areas.

\textbf{Second Level:} Captures medium-scale trends in the data with basis functions closer together.

\textbf{Third Level:} Captures fine-scale variability with tightly spaced basis functions to account for small, localized changes.

To represent a curve, we multiply each basis function by a coefficient and add them together, as shown in the following plot.The blue bars show the coefficient value at each lattice point, and the black line shows the resulting curve from scaling the basis functions with these coefficients and adding them together.

```{r IntroCurveSimulations, fig.height=3.5, fig.width=6, echo=FALSE}
info <- LKrigSetup(as.matrix(c(0,10)), alpha = 1, nlevel = 1, NC = 11, NC.buffer = 0, a.wght=2.5, LKGeometry = "LKInterval")
set.seed(483765)
coefs <- LKrig.sim(seq(0,10,,100), info, M=1, just.coefficients = TRUE)
curve <- values1 %*% coefs
plot(curve, type="l", col = 'blue', lwd = 2, x = seq(0,10,,1000), xlab = "Location",
     ylab = "Curve Value", main="Example 1-D Curve and Coefficients")
abline(v = 0:10, col="gray", lty=2)
abline(h = 0, col = "darkgray")
for (i in 1:ncol(values1)) {
  lines(gridPoints, values1[, i] * coefs[i], col = adjustcolor("black", alpha.f = 0.5))
}
barplot(coefs, col = "lightblue", width = 0.2, space = c(-0.5, rep(4, 10)), add = TRUE)

legend("topleft", legend = c("Fitted Curve", "Basis Functions", "Coefficients"),
       col = c("blue", "black", "lightblue"), lty = c(1, 1, NA), 
       lwd = c(2, 1, NA), pch = c(NA, NA, 15), pt.cex = c(1, 1, 2),
       bty = "n", inset = 0.02)

```

Now we show a visualization of simulated observed data with the true function and fitted curve in the following plot. The red dots are the observed data that represent actual data points collected (e.g., measurements of a variable like temperature or rainfall). The dashed orange curve represents the true function or "ground truth" that represents the actual relationship between the variable and the location, without noise. The fitted curve is the best approximation of the true function using the basis functions and their coefficients.

```{r, echo=FALSE}
# Created using LLM
library(gridExtra)
library(ggplot2)

# Generate grid points and centers for basis functions
grid_points <- seq(0, 10, length = 1000)
centers <- seq(0, 10, by = 2)  # Centers for the basis functions
overlap <- 2.5

# Generate true underlying function (ground truth)
set.seed(532)
true_function <- function(x) 0.8 * sin(2 * pi * x / 10) + 0.2
observed_points <- seq(0, 10, length.out = 11)  # Observed locations
observed_values <- true_function(observed_points) + rnorm(11, sd = 0.1)  # Add noise

# Calculate basis function values for observed points
basis_values <- sapply(centers, function(center) {
  phi(abs(observed_points - center) / overlap)
})

# Fit coefficients using least squares
coefficients <- solve(t(basis_values) %*% basis_values) %*% t(basis_values) %*% observed_values

# Reconstruct the fitted curve
fitted_curve <- sapply(grid_points, function(x) {
  sum(coefficients * sapply(centers, function(center) {
    phi(abs(x - center) / overlap)
  }))
})

# Visualize the process
# 1. Observed points and true function
true_curve <- data.frame(Location = grid_points, Value = true_function(grid_points))
observed_data <- data.frame(Location = observed_points, Value = observed_values)
fitted_data <- data.frame(Location = grid_points, Value = fitted_curve)

# 2. Create a list of data frames for each scaled basis function
scaled_basis_list <- lapply(seq_along(centers), function(i) {
  data.frame(
    Location = grid_points,
    Value = coefficients[i] * phi(abs(grid_points - centers[i]) / overlap),
    BasisFunction = paste0("BF_", i)
  )
})

# Combine all scaled basis function data frames into one
scaled_basis_data <- do.call(rbind, scaled_basis_list)

# Plot 1: Observed Data and Fitted Curve with Direct Labels
p1 <- ggplot() +
  geom_line(data = data.frame(Location = grid_points, Value = true_function(grid_points)),
            aes(x = Location, y = Value),
            linetype = "dashed", size = 1.2, alpha = 0.7, color = "orange") +
  geom_point(data = data.frame(Location = observed_points, Value = observed_values),
             aes(x = Location, y = Value),
             size = 3, color = "red") +
  geom_line(data = fitted_data,
            aes(x = Location, y = Value),
            size = 1.2, color = "blue") +
  labs(
    title = "Observed Data, True Function, and Fitted Curve",
    x = "Location",
    y = "Value"
  ) +
  annotate("text", x = 8, y = 0.8, label = "True Function (Dashed)", size = 4, color = "orange") +
  annotate("text", x = 8, y = 0.6, label = "Observed Data (Red Dots)", size = 4, color = "red") +
  annotate("text", x = 8, y = 0.4, label = "Fitted Curve (Blue)", size = 4, color = "blue") +
  theme_minimal()

# Plot 2: Scaled Basis Functions with Fitted Curve Overlay
p2 <- ggplot() +
  geom_line(data = scaled_basis_data,
            aes(x = Location, y = Value, group = BasisFunction), color = "gray", alpha = 0.6) +
  geom_line(data = fitted_data, aes(x = Location, y = Value), color = "blue", size = 1.2) +
  labs(
    title = "Scaled Basis Functions and Fitted Curve",
    x = "Location",
    y = "Value"
  ) +
  annotate("text", x = 8, y = 0.5, label = "Fitted Curve (Blue)", size = 4, color = "blue") +
  annotate("text", x = 2, y = -0.2, label = "Scaled Basis Functions (Gray)", size = 4, color = "darkgray") +
  theme_minimal()
# Combine the plots
grid.arrange(p1, p2, ncol = 1)
```

## Demonstration of Process for Better Intuition

As a primer for the rest of this tutorial, we'll give a brief example of fitting 2D real world data using the LatticeKrig model. The dataset we will use is from GSK pharmaceuticals that was an experiment on rats for an appetite suppressant drug. The y-axis depicts the intake of the rats over the course of the experiment and the x-axis is the span of days, with day 40 being when the appetite suppressant is removed. This is a small dataset that will allow us to more easily interrogate what is going on compared to the large datasets the LatticeKrig model is especially designed for.

```{r}
# Visualize data
s<- cbind(rat.diet$t) # Force this to be a matrix so it can be used in the LKrigSetup function
z<- rat.diet$trt

# Fit the LatticeKrig model
obj <- LatticeKrig(s, z)

plot(s, z, main = "Rat Appetite Intake Over Time", xlab = "Time (days)", ylab = "Intake",
     pch = 19, col = "blue")  # Add data points
abline(v = 40, col = "red", lty = 2)  # Add a vertical line for day 40

# Predict over a grid
sGrid <- seq(min(s), max(s), 0.5)
yHat <- predict(obj, sGrid)

# Add predicted curve to the plot
lines(sGrid, yHat, col = "green3", lwd = 2)

# Add legend for predicted curve
legend("topleft", legend = c("Observed Data", "Predicted Curve", "Day 40"),
       col = c("blue", "green3", "red"), pch = c(19, NA, NA), lty = c(NA, 1, 2), lwd = c(NA, 2, 1))
```

We can see that the fitted curve looks like a reasonably accurate fit. Next we'll take a look at the basis functions that were used to fit this curve.

```{r}
# Evaluate basis functions on grid
basis <- LKrig.basis(sGrid, obj$LKinfo)

# Plot basis functions
matplot(sGrid, basis, type = "l", lty = 1, main = "Basis Functions", xlab = "Time (days)", ylab = "Value")

# Summary of LKinfo
obj$LKinfo
```

We visualize in the plot above that the coefficients were computed and the curve was fit using three levels of basis functions, similar to the visual in the beginning of the introduction. Looking at the summary of the LKinfo below the basis function plot, we can see each level of basis functions including how many are included in each. Notice that there was a total of 68 basis functions used to fit 39 data points. That may seem excessive but that is the key motivation for using LatticeKrig over other methods for large datasets. LatticeKrig's approach ensures a smooth approximation by leveraging more basis functions than observations, which is especially beneficial for interpolating large, complex datasets. An interesting point to note is that each level of basis functions adds five basis functions on both sides on the specified amount. This is done to eliminate edge effects in this model that is inherint to spatial autoregression (SAR) models.\\

Now we'll dig in a little deeper and show how the coefficients are calculated. We'll isolate the first level of basis functions and use these to calculate the coefficients. To do this, we will first extract the basis function matrix that contains just this first level. 

```{r, fig.width=12, fig.height=6}
# Set up model parameters
LKinfo <- LKrigSetup(s,
                     a.wght = 2.01,
                     nlevel = 1, 
                     NC = 6,
                     LKGeometry = "LKInterval")

# Evaluate basis functions
look <- LKrig.basis(sGrid, LKinfo)  # Evaluate basis functions at sGrid

# Construct sparse matrix
B <- LKrigSAR(LKinfo, Level = 1)
B_dense <- spind2full(B)  # Convert sparse matrix to dense format for visualization

# Visualize basis functions and sparse matrix side by side
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), oma = c(0, 0, 0, 5))


# Plot basis functions
matplot(sGrid, look, type = "l", lty = 1, col = 1:6,
        main = "Basis Functions", xlab = "Time (days)", ylab = "Value")

# Plot sparse matrix
imagePlot(B_dense, main = "Sparse Matrix (B)", xlab = "Column Index", ylab = "Row Index")

```

When we multiply the inverse of this sparse matrix by a vector of random iid standard normal values, we will obtain a random realization of our coefficients. That is, the coefficients are calculated as
$$c = B^{-1}r, \hspace{2cm} \text{where} \hspace{2cm} r \sim N(0,I)$$

```{r, fig.width=12, fig.height=6}
# Simulate coefficients
# Basis matrix evaluated at all grid points
B<- spind2full(B) # Since B is constructed as a sparse matrix, we need to convert this into a dense matrix format to                        get a nice picture
PhiGrid<- LKrig.basis(sGrid,
                      LKinfo)

# Set up 2x2 plotting layout
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))  # 2x2 grid with appropriate margins

# Simulate and plot four realizations
set.seed(222)  # Ensure reproducibility
for (i in 1:4) {
  cSim <- solve(B) %*% rnorm(16)            # Generate coefficients
  ySim <- PhiGrid %*% cSim                 # Simulate Gaussian process
  plot(sGrid, ySim, type = "l", col = "green3", lwd = 2,
       main = paste("Simulation", i),
       xlab = "Time (days)", ylab = "Response")
}
```

Each of the above realizations of the normally distributed coefficients represents a possible fit of coefficients that are consistent with the covariance of a Gaussian Process model. Each one shows heavy correlation as a result of being a spatial autoregression (SAR). In the context of our example, we are assuming that the rat intake response over time is a realization of a Gaussian Process and the randomness of the coefficients reflects the variability and uncertainty in the process. These multiple different realizations help better visualize this concept.


## Glossary of important package functions

* \ttt{LatticeKrig}: A top level function that sets up the default spatial model, estimates some key spatial parameters, and uses the \ttt{LKrig} function for the kriging computation. \ttt{LatticeKrig} can use a minimal set of inputs and is a quick way to fit a kriging model to data.
* \ttt{LKrig}: Performs the Kriging computation and evaluates the Gaussian spatial likelihood function for a fixed LatticeKrig model. This is the main computational step in the package, and is common to all choices of geometries and models.
* \ttt{LKrigSetup}: Creates an \ttt{LKinfo} object, which is used to describe the complete spatial model for a \ttt{LatticeKrig} or \ttt{LKrig} call; especially useful for examining the effect of changing one parameter on the model.
* \ttt{surface}: Plots a fitted surface in 2D space as a color plot and adds contour lines.
* \ttt{imagePlot} or\ttt{image.plot} : Plots a dataset or fitted surface in 2D space as an image plot and adds a color bar legend.
* \ttt{predictSurface}: Takes a Kriging model and evaluates its fitted function on a grid of locations.
